Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-16 ind_files/05_unification_discussion.tex
Minor phrasing changes

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like technical gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side' under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} can thus be argued to be the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}, which resemble my equation \eqref{functional_langevin}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

>>>>

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side' under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} can thus be argued to be the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}, which resemble my equation \eqref{functional_langevin}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

### 2023-03-16 ind_files/06_nontechnical_discussion.tex
Minor phrasing changes

In Chapter \ref{chap_infD_processes}, I also postulate a `stochastic field theory' approach to modelling the evolution of quantitative traits in finite, fluctuating models. I show that this approach is consistent with known frameworks in theoretical population biology, reproducing results from quantitative genetics in the infinite population limit. The field equations I formulate are exactly complementary to a recent approach that uses spacetime white noise~\citep{week_white_2021}, and the connections are discussed in detail in section \ref{sec_disc_field_eqns}. This formulation also highlights a (somewhat heuristic) approach to the study of spacetime stochastic processes and related mathematical objects that avoids measure-theoretic tools and may be of independent interest to applied mathematicians. Broadly similar stochastic field theoretic approaches have also been proposed in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014} and models of collective motion~\citep{o_laighleis_minimal_2018}. I  conjecture that the stochastic `fundamental equations' I derive in section \ref{sec_fun_theorems} for discrete traits should also hold for quantitative traits under mild assumptions on the trait space of the quantitative traits, though no proof is attempted in this thesis. We can, however, take encouragement from the observation that~\cite{week_white_2021} have already derived some special cases of these equations\footnote{In particular, equations (21b) and (21c) in \cite{week_white_2021} are precisely the $m \to \infty$ limit of my equations for changes in the mean value of a type-level quantity (equation \ref{nD_stochastic_Price}) and changes in the variance of a type-level quantity (equation \ref{nD_stochastic_Price_variance}) respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. See section \ref{sec_disc_field_eqns} for a more detailed technical discussion.} for quantitative traits using a measure-theoretic approach under certain technical assumptions on the stochastic process under study, the most biologically important assumption being that the traits are normally distributed in the population. \cite{week_white_2021}'s approach is exactly complementary to mine, and reflects a deep duality between related mathematical structures. These two complementary approaches have long been recognized in population genetics, where the (simpler analogs of the) general approach in models of fixed total population size go by the names `branching processes approach'~(\cite{week_white_2021}'s formalism; In the classic literature, seen in work like~\cite{haldane_mathematical_1927,fisher_distribution_1931}) and `diffusion theory approach'~(My formalism; In the classic literature, seen in work like~\cite{fisher_dominance_1923,wright_evolution_1931,kimura_problems_1957}).

>>>>

In Chapter \ref{chap_infD_processes}, I also postulate a `stochastic field theory' approach to modelling the evolution of quantitative traits in finite, fluctuating models. I show that this approach is consistent with known frameworks in theoretical population biology, reproducing results from quantitative genetics in the infinite population limit. The field equations I formulate are exactly complementary to a recent approach that uses spacetime white noise~\citep{week_white_2021}, and the connections are discussed in detail in section \ref{sec_disc_field_eqns}. This formulation also highlights a (somewhat heuristic) approach to the study of spacetime stochastic processes and related mathematical objects that avoids measure-theoretic tools and may be of independent interest to applied mathematicians. Broadly similar stochastic field theoretic approaches have also been proposed in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014} and models of collective motion~\citep{o_laighleis_minimal_2018}. I  conjecture that the stochastic `fundamental equations' I derive in section \ref{sec_fun_theorems} for discrete traits should also hold for quantitative traits under mild assumptions on the trait space of the quantitative traits, though no proof is attempted in this thesis. We can, however, take encouragement from the observation that~\cite{week_white_2021} have already derived some special cases of these equations\footnote{In particular, equations (21b) and (21c) in \cite{week_white_2021} are precisely the $m \to \infty$ limit of my equations for changes in the mean value of a type-level quantity (equation \ref{nD_stochastic_Price}) and changes in the variance of a type-level quantity (equation \ref{nD_stochastic_Price_variance}) respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. See section \ref{sec_disc_field_eqns} for a more detailed technical discussion.} for quantitative traits using a measure-theoretic approach under certain technical assumptions on the stochastic process under study, the most biologically important assumption being that the traits are normally distributed in the population. My formalism and \cite{week_white_2021}'s formalism are complementary to each other, and reflect a deep duality between related mathematical structures that is well-appreciated in the applied mathematics literature. These two complementary approaches to modeling stochastic processes have long been recognized in population genetics, where the (simpler analogs of the) general approach in models of fixed total population size go by the names `branching processes approach'~(\cite{week_white_2021}'s formalism; In the classic literature, seen in work like~\cite{haldane_mathematical_1927,fisher_distribution_1931}) and `diffusion theory approach'~(My formalism; In the classic literature, seen in work like~\cite{fisher_dominance_1923,wright_evolution_1931,kimura_problems_1957}).
